---
title: "P8157_hw2_rw2844"
author: "Ryan Wei"
date: "2023-10-16"
output: 
  bookdown::pdf_document2:
    toc: no
    collapsed: no
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos = "H")
library(tidyverse)
library(caret)
library(latex2exp)
library(gstat)
library(sp)
library(nlme)
library(kableExtra)
write_matex <- function(x) {
  begin <- "$$\\begin{bmatrix}"
  end <- "\\end{bmatrix}$$"
  X <-
    apply(x, 1, function(x) {
      paste(
        paste(x, collapse = "&"),
        "\\\\"
      )
    })
  writeLines(c(begin, X, end))
}
```



## Question 1 {-}

**(a)**

```{r load_data}
load("../../datasets/MACS/MACS-VL.RData")
#length(unique(macsVL$id))
macsVL$log_vload = log(macsVL$vload)
```

First, figure \@ref(fig:macs-spaghetti) shows the CD4+ cell counts against time, with smooth curved fitted with different methods. We can see that there is a clear trend of decreasing cell counts as time since seroconversion increases. 

```{r macs-spaghetti,fig.cap="CD4+ cell counts over time, with linear, Generalized Additive Model (GAM), and lowess curve."}
sphaghetti_cd4 = 
  ggplot(macsVL, aes(x=month, y=cd4)) +
  geom_line(alpha = 0.2, aes(group = factor(id))) + geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(linetype = "Linear regression", color = "Linear regression"), linewidth = 1.5) +
  geom_smooth(method = "gam", se = FALSE, aes(linetype = "GAM", color = "GAM"), linewidth = 1.5) +
  geom_smooth(method = "loess", se = FALSE, aes(linetype = "LOWESS", color = "LOWESS"), linewidth = 1.5) +
  theme_bw() + 
  scale_linetype_manual(name = "Method", values = c(3,2,1),breaks = c( "Linear regression","GAM","LOWESS"))+
  scale_color_manual(name = "Method", values = c(3,4,2),breaks = c( "Linear regression","GAM","LOWESS")) +
  xlab("Time, months") +
  ylab("CD4+ cell counts") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )

sphaghetti_cd4
```

Next, to investigate the scientific question of interest, that is, the relationship between baseline viral load and the rate of decline of CD4+ cell counts, I plot the distribution of baseline viral load and the logarithm of baseline viral load in figure \@ref(fig:baseline-vload) and figure \@ref(fig:baseline-vload-log), respectively. From the histograms we can see that the distribution of raw baseline viral load is highly right-skewed and the absolute value of it is large with high variability. After log transformation, the distribution of $\log(\text{baseline viral load})$ is approximately normal with no outliers. Therefore, I decided to conduct the following analysis based on this transformed value.

```{r baseline-vload, fig.cap="Histogram of baseline viral load."}
macsVL_baseline = macsVL %>% group_by(id) %>% slice(1)

macsVL_baseline %>% 
  ggplot(aes(x = vload)) + 
  geom_histogram(alpha = 0.6, bins = 50) +
  geom_boxplot(position = "dodge2", outlier.color = "red")+
  theme_bw()+
  xlab(TeX("Baseline viral load")) +
  ylab(TeX("Counts")) +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )
```
```{r baseline-vload-log, fig.cap="Histogram of $\\log$(baseline viral load)."}
macsVL_baseline %>% 
  ggplot(aes(x = log(vload))) + 
  geom_histogram(alpha = 0.6, bins = 50) +
  geom_boxplot(position = "dodge2", outlier.color = "red")+
  theme_bw()+
  xlab(TeX("log(Baseline viral load)")) +
  ylab(TeX("Counts")) +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )
```


Figure \@ref(fig:scatter-baseline) and figure \@ref(fig:scatter-change-rate) show the scatterplots of the baseline CD4+ cell counts and the average CD4+ cell counts change rate (counts per month) against the baseline viral load. Here, the average CD4+ cell counts change rate is defined as the difference between the CD4+ cell counts at the end of the observation and the CD4+ cell counts at the beginning of the observation, divided by the total observation months. From both plots we can see that there is a trend for CD4+ cell counts to decline with elevated baseline viral loads, which is biologically plausible.


```{r scatter-baseline, fig.cap = "Scatterplot of baseline CD4+ cell counts versus $\\log$(baseline viral load), with linear, Generalized Additive Model (GAM), and lowess curve."}
ggplot(aes(x = log(vload), y = cd4), data = macsVL_baseline) + geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(linetype = "Linear regression", color = "Linear regression"), linewidth = 1.5) +
  geom_smooth(method = "gam", se = FALSE, aes(linetype = "GAM", color = "GAM"), linewidth = 1.5) +
  geom_smooth(method = "loess", se = FALSE, aes(linetype = "LOWESS", color = "LOWESS"), linewidth = 1.5) +
  theme_bw() + 
  scale_linetype_manual(name = "Method", values = c(3,2,1),breaks = c( "Linear regression","GAM","LOWESS"))+
  scale_color_manual(name = "Method", values = c(3,4,2),breaks = c( "Linear regression","GAM","LOWESS")) +
  xlab(TeX("log(Baseline viral load)")) +
  ylab("CD4+ cell counts") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )
```


```{r scatter-change-rate, fig.cap = "Scatterplot of average CD4+ cell counts change rate versus $\\log$(baseline viral load), with linear, Generalized Additive Model (GAM), and lowess curve."}
macsVL%>% 
  group_by(id) %>% 
  arrange(month, .by_group = T) %>% 
  mutate(cd4_change = last(cd4) - first(cd4)) %>% 
  mutate(cd4_change_rate_avg = cd4_change/(last(month)- first(month))) %>% 
  slice(1) %>% 
  ungroup() %>% 
  ggplot(aes(x = log_vload, y = cd4_change_rate_avg))+
  geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(linetype = "Linear regression", color = "Linear regression"), linewidth = 1.5) +
  geom_smooth(method = "gam", se = FALSE, aes(linetype = "GAM", color = "GAM"), linewidth = 1.5) +
  geom_smooth(method = "loess", se = FALSE, aes(linetype = "LOWESS", color = "LOWESS"), linewidth = 1.5) +
  theme_bw() + 
  scale_linetype_manual(name = "Method", values = c(3,2,1),breaks = c( "Linear regression","GAM","LOWESS"))+
  scale_color_manual(name = "Method", values = c(3,4,2),breaks = c( "Linear regression","GAM","LOWESS")) +
  xlab(TeX("log(Baseline viral load)")) +
  ylab("Average CD4+ cell counts change rate (counts/month)") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )
```

Figure \@ref(fig:cd4-remove-time) shows the residuals of the CD4+ cell counts and the logarithm of baseline viral load with time trend removed. Again, we can see that there is clearly a trend for CD4+ cell counts to decline with elevated baseline viral loads, which is biologically plausible.

```{r cd4-remove-time,fig.cap="The residuals with time trends removed for CD4+ cell counts and $\\log(\\text{Baseline viral load})$."}
fit.cd4.time <- lm(cd4 ~ month, data = macsVL)
fit.log_vload.time <- lm(log_vload ~  month, data = macsVL)

res.cd4.time <- residuals(fit.cd4.time)
res.log_vload.time <- residuals(fit.log_vload.time)

res.time.plot <-
  ggplot() +
  geom_point(aes(x = res.log_vload.time, y = res.cd4.time), alpha = 0.3) +
  geom_smooth(aes(x = res.log_vload.time, y = res.cd4.time),method = "loess", formula = y ~ x,se = FALSE, color = "red", linetype =1) +
  theme_bw() + theme(legend.position="none") +
  xlab(TeX("$\\log$(Baseline viral load)")) +
  ylab(TeX("CD4+ cell counts")) +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )
res.time.plot
```


**(b)**

The dataset has a time variable recorded in monthly intervals ranging from 0 to 48. In order to investigate the covariance structure, I decided to recode the time variable to intervals of six months, resulting in values ranging from 0 to 8. Then I used this recoded time as a covariate and CD4+ cell counts as an outcome to fit a linear marginal model and calculated the mean residual for each individual for each (categorized) time point. The following matrices provide us with the estimated variance and correlation matrix based on complete observations. The minimum number of subjects used to estimate the correlation is 85, ensuring that the estimated covariance is reliable.

$$\hat\sigma^2 = \begin{bmatrix}
346.24 \\
294.026 \\
259.777 \\
276.908 \\
301.191 \\
292.348 \\
284.811 \\
317.881 \\
313.373 \\
\end{bmatrix},\hat\rho = \begin{bmatrix}
1&0.573&0.509&0.486&0.549&0.509&0.431&0.43&0.411 \\
0.573&1&0.695&0.724&0.65&0.555&0.577&0.549&0.54 \\
0.509&0.695&1&0.742&0.642&0.604&0.617&0.565&0.486 \\
0.486&0.724&0.742&1&0.762&0.736&0.701&0.663&0.647 \\
0.549&0.65&0.642&0.762&1&0.716&0.67&0.694&0.655 \\
0.509&0.555&0.604&0.736&0.716&1&0.736&0.792&0.759 \\
0.431&0.577&0.617&0.701&0.67&0.736&1&0.806&0.781 \\
0.43&0.549&0.565&0.663&0.694&0.792&0.806&1&0.865 \\
0.411&0.54&0.486&0.647&0.655&0.759&0.781&0.865&1 \\
\end{bmatrix}.$$

From the matrices above, we can see that there isn't an obvious trend in the change of empirical standard deviations with time. The estimated correlation matrix doesn't show an obvious pattern of the correlations between observations either. Therefore, compond symmetric structure may be appropriate.


```{r vcov-eda}
macsVL$half_year = round(macsVL$month/6)
tcat <- macsVL$half_year 
#table(macsVL$half_year)
#quantile(tcat,probs = seq(0, 1, 0.1))
#table(tcat)
fit.cd4.time <- lm(cd4 ~ month, data = macsVL)
resMat.cd4 <- tapply(residuals(fit.cd4.time), list(macsVL$id, tcat), FUN=mean)
#round(resMat.fev1, 3)
res.Var <- sqrt(diag(cov(resMat.cd4, use="pairwise.complete.obs")))
res.Cov <- cor(resMat.cd4, use="pairwise.complete.obs")
nS <- matrix(NA, nrow=9, ncol=9)
for(i in 1:9){
  for(j in 1:9) nS[i,j] <- nrow(na.omit(resMat.cd4[,c(i,j)]))
}
#nS %>% as.data.frame(row.names = c(1:10))

#write_matex(as.matrix(round(res.Var,3)))
#write_matex(round(res.Cov,3))
```


Since the observation time (month) is continuous and the observed data are not balanced, we also considered the variogram that describes association among repeated observations. Figure \@ref(fig:empirical-vario) shows the variogram for observed measurements, with two components, the total variability in the data (the horizontal dashed line), and the variogram for all time lags in all individuals. We see that the difference between residuals increases as the time difference increases, which means the autocorrelations between two time points decreases as the time difference increases.

```{r empirical-vario, fig.cap = "Sample variogram of CD4+ cell counts residuals. Horizontal dashed line estimates process variance.", cache=TRUE}
library(joineR)
vgm <- variogram(indv=macsVL$id, time=macsVL$month, Y=macsVL$cd4)
#plot(vgm, smooth = TRUE, xlab="X-axis label", ylab="y-axix label")

ggplot(aes(x = vt, y = vv), data = as.data.frame(vgm[["svar"]])) + geom_point(size = 0.5,alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, aes(linetype = "LOWESS", color = "LOWESS"), linewidth = 0.8)+
  geom_hline(yintercept = vgm$sigma2, color = "black", linewidth = 0.8, linetype = 2)+
  xlab(TeX("$u_{k,ij} = |t_{ki}-t_{kj}|$")) +
  ylab(TeX("$v_{k,ij} = \\frac{1}{2}\\left(R_{kj} - R_{ki}\\right)^2$")) +
  scale_linetype_manual(name = "Method", values = c(1),breaks = c("LOWESS"))+
  scale_color_manual(name = "Method", values = c(2),breaks = c( "LOWESS"))+
  theme_bw()+
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black", size = 0.5),
  )
```


**(c)**

For this question, I first applied log transformation on baseline viral load and fitted the following model

$$[Y_{ki}] = \beta_0 + \beta_1t_{ki} + \beta_2x_k + \beta_3t_{ki}x_kE,$$
where $x_k$ is $\log(\text{Baseline viral load})$ and $t_{ki}$ is time since seroconversion
in months.

```{r gls-macs-ml}
fit.gls.ML = gls(cd4 ~ month * log_vload, method="ML", data=macsVL, corr=corCompSymm(form = ~ 1 | id))
fit.gls.REML <- gls(cd4 ~ month * log_vload, method="REML", data=macsVL, corr=corCompSymm(form = ~ 1 | id))

#summary(fit.gls.ML)
coef(summary(fit.gls.ML)) %>% 
  as.data.frame() %>% 
  mutate(term = row.names(.)) %>%
  mutate(term = c("Intercept", "Month", "$\\log$(Viral load)", "Month$\\times$(Viral load)")) %>% 
  select(term, Value, `Std.Error`, `t-value`, `p-value`) %>% 
  rename("Term" = "term", "Estimate" = "Value", "Standard Error" = "Std.Error", "T-value" = "t-value", "P-value" = "p-value") %>% 
  knitr::kable(format = "latex", digits = 3, escape = F, booktabs = TRUE, row.names = F, caption = "General least squares coefficient estimates, fit by ML.") %>% kable_styling(latex_options = "hold_position")

```

Table \@ref(tab:gls-macs-ml) and table \@ref(tab:gls-macs-reml) show the general least squares coefficient estimates fitted by maximum likelihood (ML) method and restricted maximum likelihood (REML) method. We can see that REML always gives us a estimator with larger standard error. The regression coefficient of month, $\log(\text{baseline viral load})$ and their interaction term are both significant. Table \@ref(tab:ml-reml-model-selection) shows the goodness of fit statistics of the two method, we can see that REML gives us a lower AIC as well as a lower BIC, which implies a better fit.

```{r gls-macs-reml}
#summary(fit.gls.REML)
coef(summary(fit.gls.REML)) %>% 
  as.data.frame() %>% 
  mutate(term = row.names(.)) %>%
  mutate(term = c("Intercept", "Month", "$\\log$(Viral load)", "Month$\\times$(Viral load)")) %>% 
  select(term, Value, `Std.Error`, `t-value`, `p-value`) %>% 
  rename("Term" = "term", "Estimate" = "Value", "Standard Error" = "Std.Error", "T-value" = "t-value", "P-value" = "p-value") %>% 
  knitr::kable(format = "latex", digits = 3, escape = F, booktabs = TRUE, row.names = F, caption = "General least squares coefficient estimates, fit by REML.") %>% kable_styling(latex_options = "hold_position")
```

```{r ml-reml-model-selection}
fit.gls.ML.summary = summary(fit.gls.ML)
fit.gls.REML.summary = summary(fit.gls.REML)

model.comparison.tab <-
  tibble(
    Method = c("ML", "REML"),
    AIC = c(fit.gls.ML.summary$AIC, fit.gls.REML.summary$AIC),
    BIC = c(fit.gls.ML.summary$BIC, fit.gls.REML.summary$BIC),
    logLik = c(fit.gls.ML.summary$logLik, fit.gls.REML.summary$logLik),
  )

model.comparison.tab %>% 
  knitr::kable(format = "latex", digits = 3, escape = F, booktabs = TRUE, row.names = F, caption = "Goodness of fit comparison between ML and REML.") %>% kable_styling(latex_options = "hold_position")
```

For the interpretation of the coefficients, I will pick the results from ML since both method give us a similar coefficient estimates. 

**Interpretations:**

- The CD4+ cell counts decrease by $3.09$ for each additional month after seroconversion, holding everything else the same.

- The CD4+ cell counts decrease by $35.70$ for one unit increase in $\log(\text{baseline viral load})$, holding everything else the same.

- The CD4+ cell counts decrease by $0.38$ for each additional month after seroconversion and one unit increase in $\log(\text{baseline viral load})$, holding everything else the same.



**(d)**

The quantiles of the baseline viral load in the population is shown in the table \@ref(tab:categorize-vload). Therefore, I categorized baseline viral load into four categories based on percentile of baseline viral load, corresponding to $0\%-25\%$, $25\%-50\%$, $50\%-75\%$, and $75\%-100\%$, respectively, which are Group A: $300\le \text{baseline viral load} < 7928$,  Group B: $7928\le \text{baseline viral load} < 24573$, Group C:  $24573\le \text{baseline viral load} < 91195$ and Group D: $\text{baseline viral load} \ge 91195$. Then, I fitted the following model
$$
[Y_{ki}] = \beta_0 + \beta_1t_{ki} + \sum_{j=2}^4\beta_{2j}x_k(j) + \sum_{j=2}^4\beta_{3j}t_{ki}x_k(j)
$$
```{r categorize-vload}
t(quantile(macsVL_baseline$vload)) %>% 
  as.data.frame() %>% 
  knitr::kable(format = "latex", digits = 4, escape = F, booktabs = TRUE, caption = "Quantiles of the baseline viral load.", col.names = c("$0\\%$", "$25\\%$", "$50\\%$", "$75\\%$", "$100\\%$")) %>%
  kable_styling(latex_options = "hold_position")


macsVL$vload_cat = cut(macsVL$vload, c(300, 7928, 24573, 91195, 1026657), right = FALSE, labels = c("A", "B", "C", "D"))
```

Table \@ref(tab:gls-cat-macs-reml) shows the coefficient estimates from the general least squares model fitted by REML. From the table we can see that the month is significantly associated with the CD4+ decline rate, as well as baseline viral load groups. The interactions between month and baseline viral load groups are significant except for the interaction of month with group D.

```{r gls-cat-macs-reml}
fit.gls.cat.REML = gls(cd4 ~ month * vload_cat, method="REML", data=macsVL, corr=corCompSymm(form = ~ 1 | id))
#summary(fit.gls.cat.REML)

coef(summary(fit.gls.cat.REML)) %>%
  as.data.frame() %>% 
  mutate(term = row.names(.)) %>%
  mutate(term = c("Intercept", "Month", "Viral load category B", "Viral load category C", "Viral load category D", "Month $\\times$ Viral load category B", "Month $\\times$ Viral load category C", "Month $\\times$ Viral load category D")) %>% 
  select(term, Value, `Std.Error`,`t-value`, `p-value`) %>% 
  rename("Term" = "term", "Estimate" = "Value", "Standard Error" = "Std.Error", "T-value" = "t-value", "P-value" = "p-value") %>% 
  knitr::kable(format = "latex", digits = 3, escape = F, booktabs = TRUE, row.names = F, caption = "General least squares coefficient estimates with categorized baseline viral load, fit by REML.") %>% kable_styling(latex_options = "hold_position")
```

**Interpretations:**

- The CD4+ cell counts decrease by $5.66$ on average for each additional month after seroconversion, holding everything else the same.

- The CD4+ cell counts decrease by $113.80$ on average for subjects in group B compared to subjects in group A, holding everything else the same.

- The CD4+ cell counts decrease by $133.76$ on average for subjects in group C compared to subjects in group A, holding everything else the same.

- The CD4+ cell counts decrease by $184.82$ on average for subjects in group D compared to subjects in group A, holding everything else the same.

- The CD4+ cell counts decrease by $1.66$ on average for subject in group B compared to group A, for each additional month after seroconversion, holding everything else the same.

- The CD4+ cell counts decrease by $2.11$ on average for subject in group C compared to group A, for each additional month after seroconversion, holding everything else the same.

- The CD4+ cell counts decrease by $1.21$ on average for subject in group D compared to group A, for each additional month after seroconversion, holding everything else the same. This effect is not significant.

\newpage

## Appendix: Code for this report {-}

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```